### 2018-03-12 paper/3-compute.tex
Work on S3.2

\noindent
\hs{Compute} represents an abstract set of \emph{build tasks}, for example, build
rules in software build systems or formulas in spreadsheets. It is completely
isolated from the world of compilers, file systems, calculation chains, caches,
and all other complexities of real build systems. We explain the compute type
in~\S\ref{sec-general-compute}, and then discuss several noteworthy special
cases: functorial (\S\ref{sec-compute-functor}), applicative
(\S\ref{sec-compute-applicative}) and monadic (\S\ref{sec-compute-monad}) types
of compute.

@@

We start by introducing a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets. We will use this vocabulary throughout the rest of
the paper.

@@

\subsection{Compute}\label{sec-general-compute}

@@

\noindent
A compute takes two arguments: \hs{fetch :: k -> f v} and \hs{key :: k}, and
returns either \hs{Nothing} to indicate that the \hs{key} is an input and hence
cannot be computed, or \hs{Just}~\hs{value} obtained by \hs{fetch}ing the
dependencies of the \hs{key} and performing the required computation. A compute
must be polymorphic in \hs{f} but may request certain computational
power\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
might recognise a familiar pattern. We discuss this in~\S\ref{sec-related}.} by
specifying the constraint~\hs{c}. Let us look at an example by recalling the
following spreadsheet from~\S\ref{sec-background-excel}:

@@

\subsection{Functorial compute}\label{sec-compute-functor}

@@

\subsection{Applicative compute}\label{sec-compute-applicative}
The \emph{generalised Fibonacci sequence} $F_i$ is defined as follows:

@@

\label{getdependencies} We can define the dependencies for a \hs{Compute Applicative} using the \hs{Const} functor:

\begin{minted}{haskell}
getDependencies :: Compute Applicative k v -> k -> [k]
getDependencies comp = maybe [] getConst . comp (\k -> Const [k])
\end{minted}

@@

\todo{AM}{Turn the above into a formal definition of `correctness' and add a section that provides a mathematical definition of correctness, which does not rely on materialising intermediate values.}

>>>>

\noindent
\hs{Compute} represents an abstract set of \emph{build tasks}, for example, build
rules in software build systems or formulas in spreadsheets. It is completely
isolated from the world of compilers, file systems, calculation chains, caches,
and all other complexities of real build systems. We explain the compute type
in~\S\ref{sec-general-compute}, discussing several noteworthy special cases:
functorial, applicative and monadic types of compute.

@@

We start by introducing a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets.
% We will use this vocabulary throughout the rest of the paper.

@@

\subsection{Compute}\label{sec-compute-applicative}

@@

\noindent
A compute takes two arguments: \hs{fetch :: k -> f v} and \hs{key :: k}, and
returns either \hs{Nothing} to indicate that the \hs{key} is an input and hence
cannot be computed, or \hs{Just}~\hs{value} obtained by \hs{fetch}ing the
dependencies of the \hs{key} and performing the required computation. A compute
must be polymorphic in \hs{f} but may request certain computational
power\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}
by specifying the constraint~\hs{c}. Let us look at an example by recalling the
following spreadsheet from~\S\ref{sec-background-excel}:

@@

The answer to both questions is that we can statically analyse an applicative
compute and extract all dependencies of a key using the following function:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dependencies :: Compute Applicative k v -> k -> [k]
dependencies compute = maybe [@@] getConst . compute (\k -> Const [k])
\end{minted}
\vspace{1mm}

\noindent
We can use this function to extract dependencies of \hs{add}, as shown in the
following GHCi session:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies add "A1"
[@@]
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ dependencies add "B1"
["A1", "A2"]
\end{minted}
\vspace{1mm}

\noindent
It is important to note that we extract the dependencies without providing any
values to the compute. This guarantees that no actual computation is performed
during the analysis -- we cannot execute the function \hs{(+)} if we have no
values for it! Static analysis of applicative computations is not new and has
been used to similar effect before, e.g. see~\cite{free-applicatives}.

To understand how \hs{dependencies} works, let us examine its type more closely,
expanding the type synonym \hs{Compute}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
(@\std{forall}@ f. Applicative f => (k -> f v) -> k -> Maybe (f v)) -> k -> [k]
\end{minted}
\vspace{1mm}

\noindent
Since the first argument is polymorphic in \hs{f}, we can choose any \hs{f} that
satisfies the \hs{Applicative} constraint. A good choice in this case is the
applicative functor \hs{Const} defined in the standard module
\cmd{Data.Functor.Const}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
newtype Const m a = Const { getConst :: m }

instance Functor (Const m) where
    fmap _ (Const m) = Const m

instance Monoid m => Applicative (Const m) where
    pure _              = Const mempty
    Const x <*> Const y = Const (x <> y)
\end{minted}
\vspace{1mm}

\noindent
A value of type \hs{Const}~\hs{[}\hs{k]}~\hs{v} contains no values \hs{v} but
allows us to record all invocations of the custom \hs{fetch} function that we
feed to the given compute: \hs{fetch k = Const [k]}. The compute returns a
value of type \hs{Maybe}~\hs{(Const}~\hs{[}\hs{k]}~\hs{v)}, which we unwrap
using \hs{maybe [@@] getConst} since inputs have no dependencies.

We can also execute a compute on a given store to obtain the resulting value.
This uses a very similar transformation based on the \hs{Identity} monad
defined in \cmd{Data.Functor.Identity}:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
execute :: Compute Monad k v -> (k -> v) -> k -> Maybe v
execute compute store = fmap runIdentity . compute (Identity . store)
\end{minted}
\vspace{1mm}

\noindent
Note that \hs{execute} can accept both an applicative as well as a \emph{monadic
compute} introduced in~\S\ref{sec-compute-monad}. Below we run \hs{execute} on
the above \hs{add} example. Note that \hs{execute} returns \hs{Nothing} for
inputs.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ store key = if key == "A1" then 10 else 20
@\ghci@ execute add store "A1"
Nothing
\end{minted}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
@\ghci@ execute add store "B1"
Just 30
\end{minted}
\vspace{1mm}

\subsection{Compute examples (likely to be removed)}

@@

The \emph{generalised Fibonacci sequence} $F_i$ is defined as follows:

@@

@@

\todo{AM}{Provide a formal definition of `correctness', which does not rely on materialising intermediate values.}

### 2018-03-12 paper/4-build.tex
Work on S3.2

The topological approach creates a linear ordering, which when followed, ensures each build is correct. Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort. However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{getdependencies}.

>>>>

The topological approach creates a linear ordering, which when followed, ensures each build is correct. Given a function from a key to its dependencies, and an initial key, you can compute the linear ordering by first taking the transitive closure, then doing a topological sort. However, to go from a \hs{Compute} to the dependencies, it must be the case that the constraint is \hs{Applicative}, as per \S\ref{sec-compute-applicative}.
